---
title: "Project"
author: "Caroline, Kim-Linh, Anna, Donya"
self-contained: true
code-tools: true
echo: true
message: false
warning: false
---

## Setting up R space and reading in data
```{r}
library(tidyverse)
happiness <- read_csv("hapiscore_whr.csv")
lex <- read_csv("lex.csv")
```


### Cleaning data
```{r}
lex_clean <- lex |>
  select(country, `2005`:`2022`) |>
  pivot_longer(cols = `2005`:`2022`,
               names_to = "year",
               values_to = "life_expectancy") |>
  drop_na(life_expectancy)

hap_clean <- happiness |>
  pivot_longer(cols = `2005`:`2022`,
               names_to = "year",
               values_to = "hap_score") |>
  drop_na(hap_score)
```

### Joining data
```{r}
hap_lex <-hap_clean |>
    inner_join(lex_clean)

head(hap_lex)
```

## Data and variable descriptions 

## Hypotheses about variables
We hypothesize that life expectancy will increase with happiness score. This is more of an intuitive belief, as we would expect that with a high percentage of happiness, people in the respective countries would live longer. Of course, this can vary from country to country. It is possible that more developed countries could have a higher happiness score and life expectancy rate, while less developed countries result in the opposite. 